The brightness and spatial distributions of terrestrial radio sources

### 2013-01-18 articles/Article-RFI-distribution/article-rfi-distribution.tex
Textuals

\subsection{Noise contribution} \label{sec:histogram-noise}
The full measured distribution will consist of the power-law distribution combined with that of the noise and the celestial signal. For now, we will ignore the contribution of the celestial signal, as its contribution to the amplitude distribution will be minimal when observing fields without strong sources. For example, the strongest apparent source in the NCP EoR field is around 5 Jy (Yattawata et al. 2013, \textit{submitted}). % TODO fill in details.
The standard deviation of the noise, however, is around 100~Jy on highest LOFAR resolutions, and will have a larger contribution.

@@

To understand the distribution of samples consisting of RFI + noise, one needs to consider the individual contributions of the real and imaginary components. The real and imaginary components of samples that are affected by RFI, will follow a bivariate distribution: each complex component is the sum of a complex component of RFI and a noise sample. The noise comes from a Gaussian distribution with zero mean. The RFI distribution is assumed to be a power-law distribution with yet unknown exponent.

To calculate the corresponding amplitude distribution that is formed by the combination of the real and imaginary component distributions, it is easier to perform numerical simulations, as analytic derivation is not trivial. The histogram can be numerically estimated by drawing complex real and imaginary samples from the two distributions and calculating and counting the amplitudes. A sample can be drawn from the RFI distribution by scaling and inverting the cumulative frequency function in Eq.~\eqref{eq:cumulative-rfi-distribution} and evaluating it for a uniformly distributed variable. Note however that $F_{\textrm{amplitude}\le S}(S)$ is not limited; when decreasing the amplitude $S$ towards zero, the number of sources will go to infinity. Therefore, one needs to assume that there are no samples beyond some limiting distance $r_L$ to invert the cumulative function. With this assumption, a single complex RFI contaminated sample $S_{\textrm{RFI}}$ can be sampled with
\begin{equation} \label{eq:drawing-rfi-distribution}
S_{\textrm{RFI}} \leftarrow \frac{Ig}{x_u r_L^2} e^{i 2 \pi y_u},
\end{equation}
with $x_u, y_u \sim U(0, 1)$. The result is an amplitude sample $S$, with $S \ge Ig/r_L^2$, that follows a power-law distribution. The corresponding formula for drawing a sample $S$ that is contaminated by both RFI and noise is
\begin{equation} \label{eq:drawing-combined-distribution}
S \leftarrow v_n + w_n i + \frac{Ig}{x_u r_L^2} e^{i 2 \pi y_u},
\end{equation}
with $v_n, w_n \sim N(\mu=0; \sigma)$. The final amplitude sample $S_A$ can be calculated with 
\begin{equation} \label{eq:drawing-amplitudes}
S_A \leftarrow |S| = \sqrt{\textrm{real}^2({S})+\textrm{imag}^2({S})}.
\end{equation}
An example of distribution curves for various settings of $Ig/r_L^2$ is given in Fig.~\ref{fig:rayleigh-and-rfi-combined}.

@@

For completeness we show the formula for sampling apparent source strengths that include surface propagation effects, which can be used for sampling realistic source amplitudes in simulations. This can be derived by integration, scaling and inversion of the frequency density function in Eq.~\eqref{eq:frequency-density-with-propagation}, resulting in:
\begin{equation} \label{eq:sample-with-propagation}
\tilde{S}_{\textrm{RFI}} \leftarrow \frac{\zeta I g}{x_u^{\eta/2} r_L^\eta} e^{i 2 \pi y_u}.
\end{equation}
Here, $\tilde{S}_{\textrm{RFI}}$ is a new complex RFI sample; $\zeta$ and $\eta$ are as defined below Eq.~\eqref{eq:zeta-definition}; $I$ is the average intrinsic strength; $g$ is the gain of the instrument; $0 \le x_u,y_u < 1$ are two independently drawn uniformly distributed samples; and $r_L$ is the maximum distance of visible sources.

@@

Finally, $E(S_\textrm{leak})$, which is the expected value of leaked RFI not detected by the flagger, can be calculated by replacing $S_U$ with $S_d$ in the numerator of Eq.~\eqref{eq:expected-value-rfi} and subtracting the removed number of samples from the total of number of samples. Assume that a fraction of $\eta$ samples are not detected as RFI and $1-\eta$ have been detected as RFI, then
\begin{align} \label{eq:expected-value-leaked-rfi}
E(S_\textrm{leak}) & = \frac{1}{\eta N_{LU}}\int\limits_{S_L}^{S_d} \beta S^\alpha S dS
= \frac{\left( S_d^{\alpha+2} - S_L^{\alpha+2} \right) \left( \alpha+1 \right) } {\eta \left( S_U^{\alpha+1} - S_L^{\alpha+1}\right) \left( \alpha+2 \right)}.
\end{align}
This is the average contribution that leaked RFI will have on a single sample. It has the same units as the parameters $S_L$, $S_U$ and $S_d$. Typical values for $\eta$ are $0.95$--$0.99$.

@@

\subsection{Histogram analysis}
Fig.~\ref{fig:Histograms-raw} shows the histograms with logarithmic axes for the LBA and HBA sets. In both sets, it is clear that at least one variate with a Rayleigh and one with a power-law distribution have been observed. The left part of the histogram matches the Rayleigh distribution well up to the mode of the distribution. The bulge around the mode of the LBA histogram is wider due to the larger effect of the antenna response as discussed in \S\ref{sec:parameter-variability}. As can be seen in Fig.~\ref{fig:plot-dist-per-frequency-LBA}, the Rayleigh-bulges of individual sub-bands are not that wide, but they are not aligned because of the differing noise levels. This effect is not so strong in histograms of the HBA sub-bands in Fig.~\ref{fig:plot-dist-per-frequency-HBA}, because the HBA antenna response changes less over frequency.

@@

It is to be expected that the RFI-dominated part of the distributions at different frequencies will reflect the underlying source populations. Both Figs.~\ref{fig:plot-dist-per-frequency-LBA} and \ref{fig:plot-dist-per-frequency-HBA} show that the power-law part of the distributions are very different for different sub-bands. Nevertheless, combining the data of all the sub-bands results in reasonably stable power-law distributions. The variation could be caused by the different power law exponents that source populations at different frequencies might have. It could also be caused by approximately similar underlying power-law distribution that only differ in the number of transmitters. In that case, the underlying power law might not yet be apparent, because not enough samples are combined.

@@

For the EoR project with $1$~MHz resolution, Eq.~\ref{eq:rfi-noise-level} results in $\sigma_\textrm{RFI}\approx 4.7$ mJy. However, the first EoR results of observations of one day have approximately reached the thermal noise of about $1.7$ mJy per 0.2~MHz subband (Yatawatta et al., \textit{submitted}), and the resulting images show no signs of RFI. This implies that either the upper limit is far from the actual RFI situation, or Eq.~\ref{eq:rfi-noise-level} is not applicable to most of the RFI that is observed with LOFAR. In the following section we will discuss effects that could cause a reduced contribution of RFI.

@@

Sometimes, standard data processing also attenuates RFI, mainly because of fringe stopping and the imaging of data. Fringe stopping interferometers can partly average out RFI sources. Nevertheless, stationary RFI that is averaged out by fringe stopping will leave artefacts in the field centre \citep{post-correlation-filtering}. This is not relevant when observing the North Celestial Pole --- which is one of the LOFAR EoR fields --- because no fringe stopping is applied when observing the NCP. Imaging sometimes attenuates RFI, because the Fourier transform that is involved in data imaging will localize the contribution from stationary RFI near the NCP. If RFI artefacts would show in the image of the NCP field, they can easily be detected and possibly be removed, or processing could ignore data near the pole.

Because of the above two arguments, when considering RFI it is a risk to use the NCP as one of the EoR target fields. At the same time, this field is useful for analysing the RFI coherency properties. It is also a simple field to observe with LOFAR, because it is always at high declination in the Netherlands and it does not contain bright foreground sources. Preliminary analysis of EoR NCP observations of a single night have reached the thermal noise (Yatawatta et al., \textit{submitted}), but do not show leaked RFI at the pole.

>>>>

\subsection{Noise contribution} \label{sec:histogram-noise}
The full measured distribution will consist of the power-law distribution combined with that of the noise and the celestial signal. For now, we will ignore the contribution of the celestial signal, as its contribution to the amplitude distribution will be minimal when observing fields without strong sources. For example, the strongest apparent source in the NCP EoR field is around 5 Jy (Yattawata et al. 2013, \textit{in press}). % TODO fill in details.
The standard deviation of the noise, however, is around 100~Jy on highest LOFAR resolutions, and will have a larger contribution on the histogram.

@@

To understand the distribution of samples consisting of RFI + noise, one needs to consider the individual contributions of the real and imaginary components. The real and imaginary components of samples that are affected by RFI, will follow a multi-variate distribution: each complex component is the sum of a complex component of RFI and a noise sample. The noise comes from a Gaussian distribution with zero mean. The RFI distribution is assumed to be a power-law distribution with yet unknown exponent.

To calculate the corresponding amplitude distribution that is formed by the combination of the real and imaginary component distributions, it is easier to perform numerical simulations, as analytic derivation is not trivial. The histogram can be numerically estimated by drawing complex samples from the two distributions and calculating and counting the amplitudes. A sample can be drawn from the RFI distribution by scaling and inverting the cumulative frequency function in Eq.~\eqref{eq:cumulative-rfi-distribution} and evaluating it for a uniformly distributed variable. Note however that $F_{\textrm{amplitude}\le S}(S)$ is not limited; when decreasing the amplitude $S$ towards zero, the number of sources will go to infinity. Therefore, one needs to assume that there are no samples beyond some limiting distance $r_L$ to invert the cumulative function. With this assumption, a single complex RFI contaminated sample $S_{\textrm{RFI}}$ can be sampled with
\begin{equation} \label{eq:drawing-rfi-distribution}
S_{\textrm{RFI}} \leftarrow \frac{Ig}{x_u r_L^2} e^{i 2 \pi y_u},
\end{equation}
with $x_u, y_u \sim U(0, 1)$. The result is an amplitude sample $S_{\textrm{RFI}}$, with $S_{\textrm{RFI}} \ge Ig/r_L^2$, that follows a power-law distribution. The corresponding formula for drawing a sample $S$ that is contaminated by both RFI and noise is
\begin{equation} \label{eq:drawing-combined-distribution}
S \leftarrow v_n + w_n i + \frac{Ig}{x_u r_L^2} e^{i 2 \pi y_u},
\end{equation}
with $v_n, w_n \sim N(\mu=0; \sigma)$. The final amplitude sample $S_A$ can be calculated with 
\begin{equation} \label{eq:drawing-amplitudes}
S_A \leftarrow |S| = \sqrt{\textrm{real}^2({S})+\textrm{imag}^2({S})}.
\end{equation}
An example of distribution curves for various settings of $Ig/r_L^2$ is given in Fig.~\ref{fig:rayleigh-and-rfi-combined}.

@@

For completeness we show the formula for sampling apparent source strengths that include propagation effects according to Hata's model, which can be used for sampling realistic source amplitudes in simulations. This can be derived by integration, scaling and inversion of the frequency density function in Eq.~\eqref{eq:frequency-density-with-propagation}, resulting in:
\begin{equation} \label{eq:sample-with-propagation}
\tilde{S}_{\textrm{RFI}} \leftarrow \frac{\zeta I g}{x_u^{\eta/2} r_L^\eta} e^{i 2 \pi y_u}.
\end{equation}
Here, $\tilde{S}_{\textrm{RFI}}$ is a new complex RFI sample; $\zeta$ and $\eta$ are as defined below Eq.~\eqref{eq:zeta-definition}; $I$ is the average intrinsic strength; $g$ is the gain of the instrument; $0 \le x_u,y_u < 1$ are two independently drawn uniformly distributed samples; and $r_L$ is the maximum distance of visible sources.

@@

Finally, $E(S_\textrm{leak})$, which is the expected value of leaked RFI not detected by the flagger, can be calculated by replacing $S_U$ with $S_d$ in the numerator of Eq.~\eqref{eq:expected-value-rfi} and subtracting the removed number of samples from the total of number of samples. Assume that a fraction of $\kappa$ samples are not detected as RFI and $1-\kappa$ have been detected as RFI, then
\begin{align} \label{eq:expected-value-leaked-rfi}
E(S_\textrm{leak}) & = \frac{1}{\kappa N_{LU}}\int\limits_{S_L}^{S_d} \beta S^\alpha S dS
= \frac{\left( S_d^{\alpha+2} - S_L^{\alpha+2} \right) \left( \alpha+1 \right) } {\kappa \left( S_U^{\alpha+1} - S_L^{\alpha+1}\right) \left( \alpha+2 \right)}.
\end{align}
This is the average contribution that leaked RFI will have on a single sample. It has the same units as the parameters $S_L$, $S_U$ and $S_d$. Typical values for $\kappa$ are $0.95$--$0.99$.

@@

\subsection{Histogram analysis}
Fig.~\ref{fig:Histograms-raw} shows the histograms with logarithmic axes for the LBA and HBA sets. In both sets, it is clear that at least one component with a Rayleigh and one with a power-law distribution have been observed. The left part of the histogram matches the Rayleigh distribution well up to the mode of the distribution. The bulge around the mode of the LBA histogram is wider due to the larger effect of the antenna response as discussed in \S\ref{sec:parameter-variability}. As can be seen in Fig.~\ref{fig:plot-dist-per-frequency-LBA}, the Rayleigh-bulges of individual sub-bands are not that wide, but they are not aligned because of the differing noise levels. This effect is not so strong in histograms of the HBA sub-bands in Fig.~\ref{fig:plot-dist-per-frequency-HBA}, because the HBA antenna response changes less over frequency.

@@

It is to be expected that the RFI-dominated part of the distributions at different frequencies will reflect the underlying source populations. Both Figs.~\ref{fig:plot-dist-per-frequency-LBA} and \ref{fig:plot-dist-per-frequency-HBA} show that the power-law part of the distributions are very different for different sub-bands. Nevertheless, combining the data of all the sub-bands results in reasonably stable power-law distributions. The variation could be caused by the different power-law exponents that source populations at different frequencies might have. It could also be caused by a differing number of transmitters. In that case, the underlying power law might not always be apparent, because not enough samples are combined.

@@

For the EoR project with $1$~MHz resolution, Eq.~\ref{eq:rfi-noise-level} results in $\sigma_\textrm{RFI}\approx 4.7$ mJy. However, the first EoR results of observations of one day have approximately reached the thermal noise of about $1.7$ mJy per 0.2~MHz subband (Yatawatta et al. 2013, \textit{in press}), and the resulting images show no signs of RFI. This implies that either the upper limit is far from the actual RFI situation, or Eq.~\ref{eq:rfi-noise-level} is not applicable to most of the RFI that is observed with LOFAR. In the following section we will discuss effects that could cause a reduced contribution of RFI.

@@

Some standard data processing actions can also attenuate RFI, especially fringe stopping and the imaging of data. Fringe stopping interferometers can partly average out RFI sources. Nevertheless, stationary RFI that is averaged out by fringe stopping will leave artefacts in the field centre \citep{post-correlation-filtering}. This is not relevant when observing the North Celestial Pole --- which is one of the LOFAR EoR fields --- because no fringe stopping is applied when observing the NCP. Imaging sometimes attenuates RFI, because the Fourier transform that is involved in data imaging will localize the contribution from stationary RFI near the NCP. If RFI artefacts would show in the image of the NCP field, they can easily be detected and possibly be removed, or processing could ignore data near the pole.

Because of the above two arguments, when considering RFI it is a risk to use the NCP as one of the EoR target fields. At the same time, this field is useful for analysing the RFI coherency properties. It is also a simple field to observe with LOFAR, because it is always at high declination in the Netherlands and it does not contain bright foreground sources. Preliminary analysis of EoR NCP observations of a single night have reached the thermal noise (Yatawatta et al. 2013, \textit{in press}), but do not show leaked RFI at the pole.
